Interisland movements of woodland caribou in Newfoundland

### 2020-02-19 paper/manuscript.Rmd
wolves

Predator avoidance could contribute to swimming behaviour of caribou (Miller
2002). Unlike elsewhere in the range of caribou, wolves (*Canis lupus*) are
extirpated from Newfoundland, including Fogo Island, and coyotes (*Canis latrans*) and black bears
(*Ursus americanus*) are the primary predators of caribou (Bastille-Rousseau et
al. 2016). While predation by coyotes on adult female caribou is possible, it is
relatively rare on Fogo Island (Webber & Vander Wal unpublished data).
Therefore, any effect of predation on swimming behaviour is likely related to
fine-scale encounters between coyotes and caribou which result in caribou
swimming as a flight response. Alternatively, during calving, females may swim
to small islands to give birth (Bergerud & Page 1987) or swim to islands with a
calf at heel to avoid predation. Swimming with a calf at heel has been observed
for caribou in freshwater lakes on Fogo Island (Webber & Vander Wal unpublished
data), so we anticipate it may also be possible in the ocean. We suggest future
studies assess encounter rates among caribou and coyotes to determine what
proportion of encounters result in swimming behaviour.Woodland caribou are well
known to disperse to remote locations, including islands, during calving
(Bergerud & Page 1987; Cumming & Beange 1987). Moving to remote locations is a
viable anti-predator strategy for caribou because it reduces detection and
encounter rates by predators. For instance, the use of shoreline habitat in
summer was deemed an effective anti-predator calving strategy because caribou
avoided predators by using islands and peninsulas or swimming as a flight
response (Bergerud et al. 1990). The use of water, and by association
occasionally swimming, is therefore part of the fine-scale interactions between
caribou and their predators and the use of islands appears to be an effective
anti-predator strategy. Coyotes and black bears are responsible for the majority
of mortalities for neonate caribou calves (Bastille-Rousseau et al. 2016),
although predation can still occur after the calving period (Lewis & Mahoney
2014). However, on Fogo Island, coyotes are the only predator of caribou.

>>>>

Predator avoidance could contribute to swimming behaviour of caribou (Miller
2002). Although caribou co-evolved with wolves (*Canis lupus*) throughout their range, wolves are extirpated from Newfoundland, including Fogo Island. Coyotes (*Canis latrans*)
and black bears are the primary predators of
caribou (Bastille-Rousseau et al. 2016). While predation by coyotes on adult
female caribou is possible, it is relatively rare on Fogo Island (Webber &
Vander Wal unpublished data). Therefore, any effect of predation on swimming
behaviour is likely related to fine-scale encounters between coyotes and caribou
which result in caribou swimming as a flight response. Alternatively, during
calving, females may swim to small islands to give birth (Bergerud & Page 1987)
or swim to islands with a calf at heel to avoid predation. Swimming with a calf
at heel has been observed for caribou in freshwater lakes on Fogo Island (Webber
& Vander Wal unpublished data), so we anticipate it may also be possible in the
ocean. We suggest future studies assess encounter rates among caribou and
coyotes to determine what proportion of encounters result in swimming
behaviour.Woodland caribou are well known to disperse to remote locations,
including islands, during calving (Bergerud & Page 1987; Cumming & Beange 1987).
Moving to remote locations is a viable anti-predator strategy for caribou
because it reduces detection and encounter rates by predators. For instance, the
use of shoreline habitat in summer was deemed an effective anti-predator calving
strategy because caribou avoided predators by using islands and peninsulas or
swimming as a flight response (Bergerud et al. 1990). The use of water, and by
association occasionally swimming, is therefore part of the fine-scale
interactions between caribou and their predators and the use of islands appears
to be an effective anti-predator strategy. Coyotes and black bears are
responsible for the majority of mortalities for neonate caribou calves
(Bastille-Rousseau et al. 2016), although predation can still occur after the
calving period (Lewis & Mahoney 2014). However, on Fogo Island, coyotes are the
only predator of caribou.
